An Empirical Analysis of the Role of Energy in Economic Growth

### 2014-08-09 Articles/Article_EnerEcon_2014/EconGrowth3.Rnw
Now using \Sexpr for % of data on spaghetti graphs.

% Approach
% 
% Build on articles using R to test econ growth PF, 
% question what is reported when doing these tests (shen/whalley and others who have used micEconCES). 
% 
% 
% Discussion and Insights
% 
% Does analysis differ from others in the conversation?
% Yes, differs as follows:
% \begin{itemize}
% \item{if one do these test, report on all $\delta$, $\sigma$, $\gamma$, and $\lambda$ 
% values as well as sse. 
% Good fits not evidence of best values, must report uncertainties as well.}
% \item{What motivates nesting of CES? Not well behaved $\rho$ results, driving $\sigma$s.
% Different type of CES with parsimonious factors, sigmas can be generated but not that easy (idea Randy).
% If you run all 3 nesting options, you can identify elasticities of all three factors in relation to each 
% other. Furthermore, you can determine which two factors are driving overall behavior in the aggregate function.}
% \item{necessary of different factors in PF to explain growth for countries in diff. development status}
% \end{itemize}
% 
% What do we add to the conversation?
% 
% \begin{itemize}
% 
% \item{We are able to gain rich insights 
% on what really drives growth from detailed analysis (including uncertainties) 
% by testing all functional forms across many different 
% economy types.}
% 
% \item{We were unable to gain new insights from the various energy quantifications that we tried.}
% 
% \item{We find, surprisingly, that energy is not necessary 
% to be included alongside $k$ and $l$ in all PF. Models with energy and models without energy
% all perform well on an R$^2$ basis.}
% 
% \item{We also find that 
% 
% (1) covariance of energy($e$) with both $k$ and $l$ is very high, with few exceptions (UK?).
% 
% (2) with few exceptions, $\sigma_{(kl)e}$, $\sigma_{ke}$, and $\sigma_{le}$ 
% are all very low (indicating complementarity) (verify!!!!), and 
% 
% (3) energy-only models (SF$e$ and LINEX) achieve very good fits, just like CD$e$ and CES$e$.
% (should we make this point or not?)}
% 
% \item{What does that mean? 
% It is a fallacy to think that saying energy is unnecessary 
% to include in an equation means energy is unimportant for economic growth. How to support this argument?
% 
% If you find that energy is unnecessary to include alongside capital and labor 
% to explain growth in one economy at the macro, aggregated level, 
% you should not assume that energy is unimportant as a driver for economic growth for all economies.
% For developed economies, energy is captial intensive: you need to have large capital to make use 
% of energy. In developing economies, you don't need capital to make use of energy. 
% Rather, you need labor.
% 
% We need to verify the above by comparison with our results.}
% 
% \item{Also signals that different types of countries likely 
% to react differently to energy constraints. 
% Interpret $\sigma$ here, substitution possibilities.}
% 
% \end{itemize}
% 
% More work?
% 
% 
% 
% 
% Big picture
% [insert test test]
% 
% Why do we do this? Want to link energy and economics on national level 
% and in a modeled approach,informing energy policy during energy transitions.
% Usually macro energy economic models rely on PF with energy included. 
% Key papers using CES in energy economics incl high impact such as 
% Stern, Gerlagh, Tol etc. 
% Cite in introduction met energy economists using PF approaches, specially CES.
% But, which form to choose and using which parameters? 
% Literature is inconclusive, very little actual empirical testing of such functions. 
% Mostly CES in CGE models used by Wbank and others. 
% CD adviced to be used by CBO. 
% How do we select a functional form and what are the parameters to be used? On what basis do we select?
% 
% We aim. to search for empirically robust solution, one that stands up to statistical testing.
% 
% What we find is counterintuitive for many cases: dropping energy from the PF has little impact, 
% but some notable exceptions. 
% On other hand, only including energy in a PF of LINEX form also yields good results and stats tests. 
% Is not a simple case for dropping energy, but acknowledging its key role. 
% Do we support the call for better bottom-up understanding? 
% What do we say on the limitations of macroeconomic modelling approaches altogether? 
% (Note such limitations are well known in ME literature, for example capital controversy Robinson).
% We speculate on implications for choosing PF in energy economic models that inform growth policy.
% 
% Points on paper so far
% 1. Does the relative small differences in fit between different production functions warrant 
% such a large debate on validity of one vs other? Clearly not. 
% CD been used by economists for ease of use and extreme flexibility (Miller).  
% PF generally used on instrumentalist grounds if give reasonable good description of data. 
% NK econ theory provide poor causal explanation of observed data, 
% but yet CD shows very good fit (miller, p17). 
% CES is  more general, but is a liability in forecasting because difficult 
% to get estimate of EoS that is consistent across studies (Miller, p26). 
% Aggregation is problem of both CD and CES functional forms, 
% so being careful when making strong inferences that cannot be justified.
% 
% 2. We think we can contribute by emphasising the need 
% for proper stats testing or inconsistent reporting on use of preferred functions. 
% Problems with empirical testing presented by Shen, Serrenho.
% What are our issues with multivariate OLS (Serrenho)?
% 
% 3. Are we joining the forced choice between production functions or 
% acknowledge the similar problems with aggregate prod functions in general? 
% Do we see this as sufficient evidence to doubt the top-down approach to econ growth altogether?
% 
% 4. R2 is a very crude measure and only very minor improvements 
% are achieved when others factors such as energy is added in a CD. 
% We concur. Puzzling is conclusion that such minor improvements 
% does prove the need to include energy in PF (Serrenho).
% 
% 5. We do think it is a choice between CD and CES (Miller) as 
% both have serious deficiencies after stats tests.
% 
% 6. One central issue within PF is elasticity of subs (EoS). 
% That is the \%ch in factor proportions ( k,l, etc) 
% because of one unit ch in MRTS. 
% MRTS is rate at which one factor can be substituted for another holding output constant. 
% EoS is low in Leontief cases, L shaped isoquants.
% 
% 7. Miller makes the point that even if parameters in model are stats artifacts, 
% they are still useful for forecasting. Do we agree?
% 
% 8. Is Ockham's razor relevant to what we are doing?
% 
% 9. Assumptions of CD are strongly rejected by empirical work, Miller still makes case for using CD.
% 
% 11. What about energy augmented 
% 
% 10. TZ case
% CDno e: almost 80\% l
% CD with e: 25\% energy, rest c,importance of l drops to 0
% CES no e: 20\% c ($\sigma$), 80\% l, but large range
% CES with e: fig 35 e may be NB
% LINEX: one of lowest ct parameters (low relative importance of labour eff compared to cap eff)
% Does this mean that relative small increases in cap eff can really lift the economy?
% 
% 11. ZA case
% CD no e: roughly half c, l
% CD with e: roughly half, c, e, importance of l drops to 0
% CES no e: 10\% c, 25\% l ??
% CES with e: fig 35 e may be NB
% LINEX: one of highest ct parameters (high relative importance of labour eff compared to cap eff)
% Does this mean that relative small increases in labour eff can really lift the economy?
% 
% 12. Capital starved Zambia. Very high importance. Understand CES param better.
% 
% Discussion: What is Next?
% 
% a. A comparative analysis between diff PF reveal interesting 
% enough information to warrant a more diffused approach to countries. 
% Thus, we have a case for doing this for  all countries in the world 
% (or at least those representative of countries in different groups 80/20)
% 
% b. If c is relatively more important than labour or energy 
% for growth and if capital stock is low, a situation one would expect in LDCs, 
% a relative small addition of capital stock can give relatively large returns in terms of growth.
% 
% c. Now it becomes interesting, in a 'stages of growth' approach, 
% using LT econ data it may be the case that the importance 
% of factors are different across growth stages. 
% It would need a composite type of analysis - both stock and importance.
% 
% d. Is it possible to have evidence of diff stocks and diff importances across phases of growth? 
% Maybe test this for a data rich nation that has worked through these stages, 
% we try to identify them and do partial analysis on time periods coinciding with these stages.
% 
% Should we eliminate $u$ data? Only 3 countries. Only 20 years. Doesn't add much insight. 
% Too bad, but perhaps necessary to simplify paper.
% 
% Discussions:
% 
% Counterfactuals: energy, cap, labour what ifs for future.
% 
% interpretation of type of economies, what does the data presentation say 
% on relative change in each input and change in GDP?
% 
% Forcing const elast of subs in CES thru nested structures violates 
% observation that more than two factor will not be const. elasticities Uzawa 1962.
% 
% R programming dealt with nested CES see paper discussion on Kemfert.
% 
% Next steps:
% 
% Stages of growth crowd implicitly assumes countries at similar stage of growth will
% have similar values for $\alpha$, $\beta$, $\gamma$. However, our results show
% that not to be true. Needs further exploration to cover more (all?) countries.
% 
% 
% 
% 
% **** Notes on the introduction from Martin ****
% 
% 1. Set scene through energy economic modelling papers. 
% Point out that PF approaches are alive and well (Stern, Tol, Gerlagh, World Bank, IMF), 
% short lit review as part of a longer introduction.  
% Mention competing PF approaches and parameter choice. 
% On what basis selection and how estimate parameters? 
% 
% 2. Point out the need for adequate statistical testing and economic interpretation, 
% citing some earlier critical papers on PF limitations. Take issue with 
% 
% 3. Spell out the contribution we will be making in this paper, 
% we are taking the approach that all models are wrong but some are useful. 
% We are after generating insights even after deeper assumptions of econ growth theory are very limiting.
% 
% 4. Introduce coordinates of analysis.
% 
% 
% 
% 
% 
% **** Original introduction that MKH wrote ****

@@

% \citet{Ayres:2008cl} uses Equation~\ref{eq:CDe} with exergy ($x$) and 
% useful work ($u$) quantifications for $e$ to describe historic Japanese 
% and American economies from 1900-2000. \citet{Froling:2009vo} uses an 
% energy-augmented Cobb-Douglas production function with thermal energy 
% quantification ($q$) to describe output for the global economy from 1800--1970. 
% Fr\"{o}ling's model endogenizes labor with respect to fertility 
% and mortality rates and separates thermal energy into two types: biomass 
% energy and coal energy. 
% \citet{Serrenho:2010vq} use the log transform of 
% the energy-augmented Cobb-Douglas function with thermal energy 
% quantification ($q$) to describe output for the EU-15 countries. 
% Serrenho divides thermal energy into renewables, non-renewables, 
% wastes, and imports. 
% The main argument of these papers is that 
% energy is an important driver of growth and needs to be included
% as an essential factor of production.
% 
% **** Martin: add more here about the conclusions of each paper regarding 
% the role of energy in economic growth? Add additional literature? 
% Or, adjust as you see fit. Especially, are there any confidence intervals 
% or statistical evaluation in each of these studies? ****

@@

% \citet{VanderWerf:2008tc} evalutes various nested CES production 
% functions for 12 OECD countries with respect to climate change policy. 
% \citeauthor{VanderWerf:2008tc} concludes that ($kl$)($e$) nesting structure 
% best fits historical data. \citet{Lecca:2011kr} also evaluate a variety of 
% CES production functions using a KLEM structure to determine how the various 
% options for nesting the energy factor of production affect predictive ability. 
% \citet{Stern:2012ey} also use a CES production function of the form ($kl$)($e$). 
% \citeauthor{Stern:2012ey} apply their CES function to the Swedish economy 
% from 1800-2000 and demonstrate the importance of the transition from wood to 
% fossil fuels for economic growth. \citet{Shen:2013vp} included energy 
% in nested aggregated CES production functions 
% for China and found that the best nesting structure is ($le$)($k$).
% 
% 
% **** Martin: add more here about the conclusions of each paper regarding 
% the role of energy in economic growth? Add additional literature? 
% Or, adjust as you see fit. Especially, are there any confidence 
% intervals or statistical evaluation in each of these studies? ****

@@

% \citet{Kummel:2002tx} use the 
% LINEX production function to describe economic output 
% for the economies of the U.S., Japan, and Germany. 
% \citet{Warr:2010ga} compare the LINEX production function with 
% the Cobb-Douglas production function to assess economic growth 
% for Austria, the United Kingdom, the United States, and Japan. 
% \citet{Ayres:2005kn} argued that introducing energy inputs 
% in traditional production functions does not improve its explanatory power, 
% but including useful work in a LINEX function does. 
% 
% \authNote{Martin: add more here about the conclusions of each paper 
% regarding the role of energy in economic growth? Add additional literature? 
% Or, adjust as you see fit. Especially, are there any confidence intervals 
% or statistical evaluation in each of these studies?}
% 
% \authNote{If we want to keep this quote, it should be clearer what point we are trying to
% make with it. ---rjp}
% \citet{Abramovitz:1956aa} wrote:
% 
% \begin{quote}
%   This result [that ``technological progress'' explained
%   175\% of the growth of the US economy since 1870] is surprising in the lopsided
%   importance which it appears to give to productivity increase,
%   and it should be, in a sense, sobering, if not discouraging,
%   to students of economic growth.
%   Since we know little about the causes of productivity increase,
%   the indicated importance of this element may be taken to be some sort of measure
%   of our ignorance about the causes of economic growth in the United States
%   and some sort of indication of where
%   we need to concentrate our attention.
% \end{quote}

@@

Figure~\ref{fig:CD_GDP_Spaghetti_Graph} compares predictions 
from the Cobb-Douglas models 
(without and with energy) to historical data.
Historical data are shown as a black line. 
The fit to historical data is shown as a white line.
The gray band encompasses 95\% of the fits to resampled data.

@@

Figure~\ref{fig:CES_Spaghetti_Graph} compares CES model predictions 
to historical data for all nestings. 
As with the Cobb-Douglas results, 
historical data are shown as a black line, 
the fit to historical data is shown as a white line, and 
the gray band encompasses 95\% of the fits to resampled data.

@@

% Figure~\ref{fig:CES_Spaghetti_Graph_CN} shows details of the resample fits for China with the (kl)e nesting.
% Again, historical data are shown as a black line. 
% The fit to historical data is shown as a white line.
% The gray band encompasses 95\% of the fits to resampled data.
% 
% <<CES_Spaghetti_Graph_CN, fig.pos="H", fig.align='center', fig.width=maxWidth, fig.height=maxHeight, fig.cap="CES fitted models for China with the (kl)e nesting.">>=
% spaghettiPlot(data=subset(cesFittedData, Country=="CN" & nestStrParen=="(iK + iL) + (iQp)"), level=spaghettiGraphLevel) +
%     labs(y="Indexed GDP (1980 = 1 or 1991 = 1")
% @

@@

Figure~\ref{fig:LINEX_Spaghetti_Graph} compares LINEX model predictions to historical data.
Again, historical data are shown as a black line, 
the fit to historical data is shown as a white linem and 
the gray band encompasses 95\% of the fits to resampled data.

>>>>

@@

@@

@@

@@

Figure~\ref{fig:CD_GDP_Spaghetti_Graph} compares predictions 
from the Cobb-Douglas models 
(without and with energy) to historical data.
Historical data are shown as a black line. 
The fit to historical data is shown as a white line.
The gray band encompasses \Sexpr{spaghettiGraphLevel * 100.0}\% of the fits to resampled data.

@@

Figure~\ref{fig:CES_Spaghetti_Graph} compares CES model predictions 
to historical data for all nestings. 
As with the Cobb-Douglas results, 
historical data are shown as a black line, 
the fit to historical data is shown as a white line, and 
the gray band encompasses \Sexpr{spaghettiGraphLevel * 100.0}\% of the fits to resampled data.

@@

@@

Figure~\ref{fig:LINEX_Spaghetti_Graph} compares LINEX model predictions to historical data.
Again, historical data are shown as a black line, 
the fit to historical data is shown as a white linem and 
the gray band encompasses \Sexpr{spaghettiGraphLevel * 100.0}\% of the fits to resampled data.
